Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-11-13 paper/paper.tex
discussion

Forward and cross-decomposition (CCA, PLS) models are regularly used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g. \citep{naselaris2011encoding}). These approaches typically lead to multiple coefficients for each features (i.e. one per dimension of $Y$ or one per component respectively). Furthermore, these  coefficients can be difficult to summarize into a single causal estimate. By contrast, B2B quickly (Fig. \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features respectively.

@@

Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects. First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$. By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions. Second, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability. Third, CCA does not typically optimize two distinct regularization parameters. Finally, CCA does not use different data splits to estimate $G$ and $H$. Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and \ref{fig:auc_plots}). Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal discovery.

>>>>

Forward and cross-decomposition models have been used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g. \citep{naselaris2011encoding}). These approaches typically lead to multiple coefficients for each features (i.e. one per dimension of $Y$ or one per component respectively). Furthermore, these coefficients can be difficult to summarize into a single causal estimate. By contrast, B2B quickly (Fig. \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features respectively.

@@

Finally, CCA has been used in neuroimaging for a variety of purposes such as denoising and subject alignment \citep{cca_hotelling, de2019multiway}. While CCA relates to B2B, these two methods diverge in several ways.
First, CCA and B2B have different objectives: CCA aims to find the potentially numerous and poorly interpretable components where X and Y are maximally correlated, whereas B2B aims to recover the causal factors from X to Y.
Second, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$. By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions.
Third, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability and potentially dilute the impact of each feature onto multiple components. In contrast B2B assesses each feature $X^j$ on a single $Y$ component specifically selected to maximize signal-to-noise ratio of that feature $j$. Fourth, and unlike B2B, CCA does not separately optimize two distinct regularization parameters for $G$ and $H$.
Finally, CCA does not use different data splits to estimate $G$ and $H$.
Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and \ref{fig:auc_plots}).
